Hybrid platforms and bargaining power

### 2024-08-19 src/paper/application.tex
Fix reference
Rename motivation to introduction
Grammarly intro

\section{Motivation}

Although market structures resembling various aspects of what we call platforms have existed for some time, the term has only gained widespread usage in the 2000s, with the advent of digitization, and a number of behemoth technology companies playing matchmakers on the internet.
Since then, platforms have become a buzzword in the business world, the subject of increasing regulatory scrutiny, and also a topic of intense academic interest.
Platforms have a number of unique features that make them different from more traditional market structures.
The most prominent of those are multi-sidedness and network effects, which most of the early literature \parencite[for an overview, see][]{rochet2006two} and policy debate \parencite[e.g.][]{fletcher2021consumer,calvano2021market} has focused on.
However, in the more recent years, another important and potentially concerning aspect has been getting more attention: the hybrid operation of certain platforms.
Essentially, this means that platforms act as an intermediary (deciding the rules) and a market participant (competing with other entrants) at the same time.

While hybrid platforms are not a completely new phenomenon, they are becoming more and more common in various industries.
Perhaps the most prominent example is Amazon, which, on the one hand sells its own products, and at the same time hosts ax large number of third-party sellers.
However, examples abound in other industries, as well.
For example, each of the largest digital distribution platforms for computer and smartphone applications (Google Play, Apple App Store and Microsoft Store) sell their own applications besides their third party offerings.
In the video game industry, platform owners studios and publishing games under their own brand is the standard.
Many video streaming services (e.g. Netflix, Amazon Prime Video, Hulu) offer a mix of their own and licensed content.
One can also find examples of similar behavior outside the platform setting, too.
For example, a number of car manufacturers are planning to sell directly to consumers in addition to selling through their dealership networks.

As deciding the rules of the game and competing in it at the same time gives rise to some obvious concerns, policymakers have been paying increasingly close attention to these platforms in the recent years.
Various pieces of legislation have been proposed to regulate, or even ban e-commerce platforms from selling their own products \parencite[][]{phartiyal_2019,reynolds_2022,eu-2022}.
Furthermore, one of the most prominent recent antitrust cases, Microsoft acquiring the video game publisher Activision/Blizzard for a sum of 69 billion USD, is also a case of a platform becoming increasingly hybrid.
This merger not only increased market concentration on the software side of the video game industry, but might also have further implications for the competition between Microsoft as a platform owner and the other game developers.
The deal was under investigation by multiple antitrust authorities, including the US Federal Trade Commission, the European Commission, and the UK Competition and Markets Authority \parencite{livni_merced_2023}.

In parallel to the increasing regulatory interest, the academic literature on the topic of hybrid platforms has also been steadily growing.
\textcite{hagiu2022should} investigate how practices, such as self-preferencing and steering can distort competition on hybrid platforms.
They argue that -- while such practices are problematic and can have negative welfare effects -- with proper regulation the platform selling its own products can be beneficial for consumers.
In contrast to this positive result, \textcite[]{anderson2021hybrid} find that even in the absence of such behavior, allowing hybrid operation might have negative consequences due to platforms having an incentive to exclude competitors from the market.
\textcite[]{gutierrez2021welfare} shows that general conclusions are hard to draw, and welfare effects are platform and product specific.
For example, using a mechanism design approach, \textcite{kang2022contracting} demonstrate that welfare consequences greatly depend on whether the platform faces competition in the upstream market.

This paper aims to contribute to this discussion by examining an important, but so far underexplored aspect of platforms: the bargaining power disparity between them and the other market participants, and how this disparity is affected by the platforms' hybrid operation.
I propose an analytically tractable model of a which captures many important aspects of bargaining between one large and a continuum of small players: namely a hybrid platform facing a continuum of potential entrants.
I show that, in the presence of bargaining, hybrid operation can indeed have negative welfare consequences for consumers.
This result is particularly striking in this otherwise distortion-free\footnote{
    In the benchmark model without bargaining, the platform's hybrid operation does not affect consumer welfare.
} setting. 
The intuition is that hybrid operation increases the platform's bargaining power against the entrant sellers.
This in turn leads to a higher entry fee, fewer entrants, then results in lower consumer surplus.
These observations constitute a so-far overlooked aspect of hybrid platforms that policymakers should be aware of.
Due to the relative generality of this model, these results are applicable not only to hybrid platform, but to other similar settings, such as upstream producers having their own downstream outlets, or retailers with private labels, as well.

Many of the results are not tied to the specifics of the model, but are more general instead.
They do not depend on a specific demand structure and investment cost assumptions, and there is even some flexibility to the assumptions about bargaining outcomes.
I show that, even in such stylized setting, some interesting observations can be made about the effects of hybrid operation. 
In particular, when the platform's and the potential entrants' products are substitutes, increasing the platform's product variety will lead to a decrease in the number of entrants.
I also give sufficient conditions for this decrease in entry being so large as to cause a decrease in total profits, as well.

The closest model to the one proposed in this paper is \textcite{anderson2021hybrid}.
The key difference between the two boils down to how the entry fee is decided.
In \textcite{anderson2021hybrid}, the platform sets a percentage entry fee, and the entrants decide whether to enter or not.
In contrast to this, I assume that (1) platforms charge a lump-sum entry fee and that (2) this entry fee is the result of a negotiation process\footnote{
    The bargaining process is modeled using a solution concept from cooperative game theory, namely the Shapley-value.
    This allows for a tractable analysis while at the same time capturing many of the important features of bargaining.
    Examples of this approach in the industrial organization literature include \textcite{montez2007downstream}, as well as \textcite{hart1990property}, \textcite{levy1997individual}, \textcite{inderst2003bargaining} and \textcite{brugemann2019intra} among others.
}
between the platform and the entrants.
The second point is a novelty not only compared to \textcite{anderson2021hybrid}, but also to the existing literature on hybrid platforms.
Furthermore, in contrast to \textcite{hagiu2022should}, this model is free from any distortive behavior, such as self-preferencing or steering, and the negative results are purely driven by changes in the platform's bargaining position.

The set of more general results in this paper go beyond the aforementioned studies in the sense that they illuminate the effects of hybridness as a more abstract phenomenon, and thus also highlight the similarities between the hybrid platform literature and adjacent ones.
Certain structures, such as vertical relationships, or even traditional retail stores share a number of features with platforms (notably, the presence of a dominant, indispensable entity), and can be modeled in much the same way.
Therefore, the methodology and results in this paper are also of interest to researchers studying questions such as retailers having their own private labels \parencite{steiner2004nature} or vertical integration in upstream-downstream markets \parencite{hart1990vertical,aghion2006vertical}.
\textcite{de2005vertical} and \textcite{montez2007downstream}, in which bargaining between the upstream and downstream firms take the center stage, are particularly closely related.
Among empirical studies in related settings, \textcite{ho2017insurer} looks at the effect of insurer competition on health care prices, while \textcite{crawford2018welfare} explores the effects of vertical integration on television markets.

@@

Finally, this paper also belongs to the not too large set of models using concepts from cooperative game theory in an industrial organization setting.
Such examples include the aforementioned \textcite{montez2007downstream}, as well as \textcite{hart1990property}, \textcite{levy1997individual}, \textcite{inderst2003bargaining} and \textcite{brugemann2019intra}, among others.
In contrast to the majority of those, which focus on games with a finite number of players, I utilize an oceanic game (a continuum of small players instead of a finite number), and demonstrate that this can considerably simplify the analysis in certain cases.
Therefore, the modeling approach presented in this paper may be of use in other settings as well.

@@

\Cref{prop:hybrid_vs_retail} is then a special case of this result.
\begin{proof}[Proof of \Cref{hybrid_vs_retail}]
    $\Pi$ satisfies \Cref{ass:additive_profit,ass:monotone_profits,ass:monotone_profits}.
    The result is an immediate corollary of \Cref{prop:hybrid_vs_retail_general} with $N_F=0$ and $N'F=N_F(N_P)$.
\end{proof}

>>>>

\section{Introduction}

Although market structures resembling various aspects of what we call platforms have existed for some time, the term gained widespread usage only in the 2000s, with the advent of digitization and a number of behemoth technology companies playing matchmakers on the Internet.
Since then, the concept of a platform has become prominent in the business world, drawn increasing regulatory attention, and become a topic of intense academic interest.
Platforms have several unique features that differentiate them from more traditional market structures.
The most prominent of those are multi-sidedness and network effects, which most of the early literature \parencite[for an overview, see][]{rochet2006two} and policy debate \parencite[e.g.][]{fletcher2021consumer,calvano2021market} has focused on.
However, in recent years, another important and potentially concerning aspect has been getting more attention: the hybrid operation of certain platforms.
Such platforms act as intermediaries (deciding the rules) and market participants (competing with other entrants) at the same time.

While hybrid platforms are not an entirely new phenomenon, they are becoming increasingly common in various industries.
Perhaps the most prominent example is Amazon, which sells its own products while simultaneously hosting a large number of third-party sellers.
However, examples abound in other industries, as well.
For example, each of the largest digital distribution platforms for computer and smartphone applications (Google Play, Apple App Store, and Microsoft Store) sells its own applications in addition to third-party offerings.
In the video game industry, platforms owning studios and publishing games under their own brand is the standard.
Many video streaming services (e.g. Netflix, Amazon Prime Video, Hulu) offer a mix of their own and licensed content.
One can find examples of similar behavior outside the platform setting, too.
For example, several car manufacturers are planning to sell directly to consumers in addition to selling through their dealership networks.

As deciding the rules of the game and competing in it at the same time gives rise to some obvious concerns, policymakers have been paying increasingly close attention to these platforms in recent years.
Various pieces of legislation have been proposed to regulate or even ban e-commerce platforms from selling their own products \parencite[][]{phartiyal_2019,reynolds_2022,eu-2022}.
Furthermore, one of the most prominent recent antitrust cases, Microsoft acquiring the video game publisher Activision/Blizzard for 69 billion USD, is also a case of a platform becoming increasingly hybrid.
This merger not only increased market concentration on the publishing side of the video game industry, but also had implications for the competition between Microsoft as a platform owner and the other game publishers.
The deal was under investigation by multiple antitrust authorities, including the US Federal Trade Commission, the European Commission, and the UK Competition and Markets Authority \parencite{livni_merced_2023}.

Parallel to the increasing regulatory interest, the academic literature on hybrid platforms has also been steadily growing.
\textcite{hagiu2022should} investigate how practices, such as self-preferencing and steering, can distort competition on hybrid platforms.
They argue that, while such practices are problematic and can have negative welfare effects, with proper regulation the platform selling its own products can be beneficial for consumers.
In contrast to this positive result, \textcite[]{anderson2021hybrid} find that even in the absence of such behavior, allowing hybrid operation might have negative consequences due to platforms' incentive to exclude competitors from the market.
\textcite[]{gutierrez2021welfare} shows that general conclusions are hard to draw, and welfare effects are platform and product-specific.
For example, using a mechanism design approach, \textcite{kang2022contracting} demonstrates that welfare consequences greatly depend on whether the platform faces competition in the upstream market.

This paper aims to contribute to this discussion by examining an important but underexplored aspect of platforms: the bargaining power disparity between the latter and the other market participants, and how this disparity is affected by the platforms' hybrid operation.
I propose an analytically tractable that captures many important aspects of bargaining between one large and a continuum of small players: namely, a hybrid platform facing a continuum of potential entrants.
I show that hybrid operation can indeed have negative welfare consequences for consumers in the presence of bargaining.
This result is particularly striking in this otherwise distortion-free\footnote{
    The platform's hybrid operation does not affect consumer welfare in the benchmark model without bargaining.
} setting. 
The intuition is that hybrid operation increases the platform's bargaining power against the entrant sellers.
This, in turn, leads to a higher entry fee and fewer entrants, resulting in lower consumer surplus.
These observations constitute a so far overlooked aspect of hybrid platforms that policymakers should be aware of.
Due to this model's generality, these results are applicable not only to hybrid platforms, but also to many other settings, such as upstream producers having their own downstream outlets or retailers with private labels.

Many of the results are not tied to the specifics of the model but are more general.
They do not depend on a specific demand structure, and there is even some flexibility with regard to the assumptions about bargaining outcomes.
I show that, even in such a stylized setting, some interesting observations can be made about the effects of the hybrid operation. 
In particular, when the platform's and the potential entrants' products are substitutes, increasing the platform's product variety will lead to a decrease in the number of entrants.
I also give sufficient conditions for this decrease in entry being so large as to cause a decrease in total profits.

The closest model to the one proposed in this paper is \textcite{anderson2021hybrid}.
The key difference between the two boils down to how the entry fee is determined.
In \textcite{anderson2021hybrid}, the platform sets a percentage entry fee, and the entrants decide whether or not to enter.
In contrast to this, I assume that (1) platforms charge a lump-sum entry fee and that (2) this entry fee is the result of a negotiation process\footnote{
    The bargaining process is modeled using a solution concept from cooperative game theory, namely the Shapley value.
    This allows for a tractable analysis while at the same time capturing many of the essential features of bargaining.
    Examples of this approach in the industrial organization literature include \textcite{montez2007downstream}, as well as \textcite{hart1990property}, \textcite{levy1997individual}, \textcite{inderst2003bargaining} and \textcite{brugemann2019intra} among others.
}
between the platform and the entrants.
The second point is a novelty not only compared to \textcite{anderson2021hybrid}, but also to the existing literature on hybrid platforms.
Furthermore, in contrast to \textcite{hagiu2022should}, this model is free from any distortive behavior, such as self-preferencing or steering, and the negative results are purely driven by changes in the platform's bargaining position.

The set of more general results in this paper goes beyond the aforementioned studies in the sense that they illuminate the effects of hybrid operation as a more abstract phenomenon, and thus also highlight the similarities between the hybrid platform literature and adjacent ones.
Certain structures, such as vertical relationships or even traditional retail stores, share several features with platforms (notably, the presence of a dominant, indispensable entity) and can be modeled in much the same way.
Therefore, the methodology and results in this paper are also of interest to researchers studying questions such as retailers having their own private labels \parencite{steiner2004nature} or vertical integration in upstream-downstream markets \parencite{hart1990vertical,aghion2006vertical}.
\textcite{de2005vertical} and \textcite{montez2007downstream}, in which bargaining between the upstream and downstream firms take center stage, are particularly closely related.
Among empirical studies in related settings, \textcite{ho2017insurer} looks at the effect of insurer competition on health care prices, while \textcite{crawford2018welfare} explores the effects of vertical integration on television markets.

@@

Finally, this paper also belongs to the relatively small set of models that use concepts from cooperative game theory in an industrial organization setting.
Such examples include the aforementioned \textcite{montez2007downstream}, as well as \textcite{hart1990property}, \textcite{levy1997individual}, \textcite{inderst2003bargaining} and \textcite{brugemann2019intra}, among others.
In contrast to the majority of those, which focus on games with a finite number of players, I utilize an oceanic game (a continuum of small players instead of a finite number), demonstrating that this can considerably simplify the analysis in certain cases.
Therefore, the modeling approach presented in this paper may also be useful in other settings.

@@

\Cref{prop:hybrid_vs_retail} is then a special case of this result.
\begin{proof}[Proof of \Cref{prop:hybrid_vs_retail}]
    $\Pi$ satisfies \Cref{ass:additive_profit,ass:monotone_profits,ass:monotone_profits}.
    The result is an immediate corollary of \Cref{prop:hybrid_vs_retail_general} with $N_F=0$ and $N'F=N_F(N_P)$.
\end{proof}
